---
title: "Quarto and gt tables to reveal.js slides"
author: "Jeremy Allen"
date: "`r as.Date(Sys.time())`"
format:
  revealjs:
    scrollable: true
    theme: ["theme/styles.scss"]
    slide-number: c/t
    logo: "https://www.rstudio.com/wp-content/uploads/2018/10/RStudio-Logo-Flat.png"
    footer: "[https://github.com/jeremy-allen/quarto-demo](https://github.com/jeremy-allen/quarto-demo)"
    code-copy: true
    center-title-slide: false
    include-in-header: heading-meta.html
    code-link: true
    code-overflow: wrap
    highlight-style: a11y
    height: 1080
    width: 1920
execute: 
  eval: true
  echo: true
editor_options: 
  chunk_output_type: console
---

# Penguin Anatomy Over Time and Islands: a demo of `Quarto`, `gt`, and `gtExtras`

. . .

Quarto®
: Quarto is an open-source scientific and technical publishing system built on Pandoc.  

. . .

gt
: With the gt package, anyone can make wonderful-looking tables using the R programming language, and a cohesive set of table parts, such as the table header, stub, column labels and spanner column labels, table body, and the table footer.

. . .

gtExtras
: The goal of gtExtras is to provide some additional helper functions to assist in creating beautiful tables with gt.  

## Load packages and prep data

```{r}
#| message: false
#| warning: false
#| echo: true

library(tidyverse)
library(palmerpenguins)
library(gt)
library(gtExtras)

# make a table of our data
# but let's summarize by year, so first make a year column
dat <- penguins %>%
  dplyr::arrange(year) %>% 
  mutate(year = as.factor(year)) %>% 
  group_by(species, island, year) %>% # order here will influence table output
  summarise(across(bill_length_mm:body_mass_g, ~ mean(., na.rm = TRUE))) %>%
  select(species, island, year, everything()) %>% 
  dplyr::arrange(species, island, year) %>% 
  ungroup() # experiment with and without
```

## Build the table with gt, include some gtExtras functions for colors

`gt`'s default is to produce an html table  
Let's first group our data by penguin species

```{r}
#| echo: true
#| message: false
#| warning: false
#| output-location: column
#| code-line-numbers: "|3"

# table using gt package
dat %>% 
  dplyr::group_by(species) %>% # add a grouping variable and drop stub info
  gt()

```

## Headers and spanners

```{r}
#| echo: true
#| message: false
#| warning: false
#| output-location: column
#| code-line-numbers: "|5-8|9-12"

# table using gt package
dat %>% 
  dplyr::group_by(species) %>% # add a grouping variable and drop stub info
  gt() %>% 
  tab_header(
    title = md("**Penguin Anatomical Changes**"), #wrapped in md() for markdown
    subtitle = "By species, island, and year from 2007-2009"
  ) %>%
  tab_spanner( # tab spanner!
    label = "mean anatomical measures",
    columns = c(bill_length_mm:body_mass_g)
  )
```

## Formatting numbers

```{r}
#| echo: true
#| message: false
#| warning: false
#| output-location: column
#| code-line-numbers: "|13-16"

# table using gt package
dat %>% 
  dplyr::group_by(species) %>% # add a grouping variable and drop stub info
  gt() %>% 
  tab_header(
    title = md("**Penguin Anatomical Changes**"), #wrapped in md() for markdown
    subtitle = "By species, island, and year from 2007-2009"
  ) %>%
  tab_spanner( # tab spanner!
    label = "mean anatomical measures",
    columns = c(bill_length_mm:body_mass_g)
  ) %>% 
  fmt_number(
    columns = bill_length_mm:body_mass_g,
    decimals = 1
  )
```

## Sources, footnotes, and references

```{r}
#| echo: true
#| message: false
#| warning: false
#| output-location: column
#| code-line-numbers: "|17-19|20-26|27-30"

# table using gt package
dat %>% 
  dplyr::group_by(species) %>% # add a grouping variable and drop stub info
  gt() %>% 
  tab_header(
    title = md("**Penguin Anatomical Changes**"), #wrapped in md() for markdown
    subtitle = "By species, island, and year from 2007-2009"
  ) %>%
  tab_spanner( # tab spanner!
    label = "mean anatomical measures",
    columns = c(bill_length_mm:body_mass_g)
  ) %>% 
  fmt_number(
    columns = bill_length_mm:body_mass_g,
    decimals = 1
  ) %>% 
  tab_source_note(
    source_note = "Source: Very cold nights."
  ) %>%
  tab_source_note(
    # hitting return to start new lines so the string will stay inside the PDF
    # but it will be a continuous string when it renders as the footnote
    source_note = md("Reference: Horst AM, Hill AP, Gorman KB (2020). 
    palmerpenguins: Palmer Archipelago (Antarctica) penguin data. R package
    version 0.1.0. https://allisonhorst.github.io/palmerpenguins/")
  ) %>% 
  tab_footnote(
    footnote = "Only found on a single island.",
    locations = cells_row_groups(groups = c("Chinstrap", "Gentoo"))
  )
```

## `gtExtras` for heatmap-like colors

```{r}
#| echo: true
#| message: false
#| warning: false
#| output-location: column
#| code-line-numbers: "|31-35"

# table using gt package
dat %>% 
  dplyr::group_by(species) %>% # add a grouping variable and drop stub info
  gt() %>% 
  tab_header(
    title = md("**Penguin Anatomical Changes**"), #wrapped in md() for markdown
    subtitle = "By species, island, and year from 2007-2009"
  ) %>%
  tab_spanner( # tab spanner!
    label = "mean anatomical measures",
    columns = c(bill_length_mm:body_mass_g)
  ) %>% 
  fmt_number(
    columns = bill_length_mm:body_mass_g,
    decimals = 1
  ) %>% 
  tab_source_note(
    source_note = "Source: Very cold nights."
  ) %>%
  tab_source_note(
    # hitting return to start new lines so the string will stay inside the PDF
    # but it will be a continuous string when it renders as the footnote
    source_note = md("Reference: Horst AM, Hill AP, Gorman KB (2020). 
    palmerpenguins: Palmer Archipelago (Antarctica) penguin data. R package
    version 0.1.0. https://allisonhorst.github.io/palmerpenguins/")
  ) %>% 
  tab_footnote(
    footnote = "Only found on a single island.",
    locations = cells_row_groups(groups = c("Chinstrap", "Gentoo"))
  ) %>%
  # trim gives smaller range of colors
  # so the green and purples are not as dark
  gt_hulk_col_numeric(bill_length_mm, trim = TRUE) %>%  # gtExtras!
  gt_hulk_col_numeric(bill_depth_mm, trim = TRUE) %>%  # gtExtras!
  gt_hulk_col_numeric(flipper_length_mm, trim = TRUE)  # gtExtras!
  
```

## Change column labels and row-group lables

```{r}
#| echo: true
#| message: false
#| warning: false
#| output-location: column
#| code-line-numbers: "|36-43|44-46"

# table using gt package
dat %>% 
  dplyr::group_by(species) %>% # add a grouping variable and drop stub info
  gt() %>% 
  tab_header(
    title = md("**Penguin Anatomical Changes**"), #wrapped in md() for markdown
    subtitle = "By species, island, and year from 2007-2009"
  ) %>%
  tab_spanner( # tab spanner!
    label = "mean anatomical measures",
    columns = c(bill_length_mm:body_mass_g)
  ) %>% 
  fmt_number(
    columns = bill_length_mm:body_mass_g,
    decimals = 1
  ) %>% 
  tab_source_note(
    source_note = "Source: Very cold nights."
  ) %>%
  tab_source_note(
    # hitting return to start new lines so the string will stay inside the PDF
    # but it will be a continuous string when it renders as the footnote
    source_note = md("Reference: Horst AM, Hill AP, Gorman KB (2020). 
    palmerpenguins: Palmer Archipelago (Antarctica) penguin data. R package
    version 0.1.0. https://allisonhorst.github.io/palmerpenguins/")
  ) %>% 
  tab_footnote(
    footnote = "Only found on a single island.",
    locations = cells_row_groups(groups = c("Chinstrap", "Gentoo"))
  ) %>%
  # trim gives smaller range of colors
  # so the green and purples are not as dark
  gt_hulk_col_numeric(bill_length_mm, trim = TRUE) %>%  # gtExtras!
  gt_hulk_col_numeric(bill_depth_mm, trim = TRUE) %>%  # gtExtras!
  gt_hulk_col_numeric(flipper_length_mm, trim = TRUE) %>%  # gtExtras!
  cols_label( # new column labels!
    bill_length_mm = html("Bill Length<br>mm"),
    bill_depth_mm = html("Bill Depth<br>mm"),
    flipper_length_mm = html("Flipper Length<br>mm"),
    body_mass_g = html("Body Mass<br>g"),
    island = "Island",
    year = "Year"
  ) %>% 
  tab_options(
    row_group.font.weight = "800"
  )
```

# Your turn! To get started:
see [quarto.org](https://quarto.org/)  
see [gt](https://gt.rstudio.com/index.html)  
see [gtExtras](https://jthomasmock.github.io/gtExtras/) 

